---
title: "Music Cognition and Memory"
format:
  dashboard:
    theme:
      - superhero
      - styles.scss
---

```{r setup}
#| include: false
library(tidyverse)
library(ggplot2)
library(dplyr)

df <- read.csv("qualtrics data.csv")

df <- df %>%
  rename(
    With.music.or.not = Music.Background,
    Memory.Score = Score,
    Total.Gold.MSI.Score = TOTAL.MSI.SCORE
  ) %>%
  mutate(With.music.or.not = factor(With.music.or.not, levels = c("No", "Yes")))

```


# Introduction 

<div style="color:#000;">

For our research on everyday music listening we decided to investigate the question

How does passive music listening affect performance for people of different levels of music cognition?

We think that it is worth investigating the possibility that people with a high knowledge of music theory might get more or less distracted/affected by the presence of background music when performing a memory task.

To do this we presented all of our subjects with the exact same memory task, except half of them (selected randomly) had to perform the tasks.
</div>


## Row 1
### Column 1 {width="60%"}
```{r}
#| title: Memory Test Scores Dependent on Background Music and Music Proficiency
#| echo: false

ggplot(df, aes(
  x = Total.Gold.MSI.Score,
  y = Memory.Score,
  color = With.music.or.not
)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(
    values = c(
      "No" = "#1f77b4",
      "Yes" = "#ff7f0e"
    )
  ) +
  labs(
    title = "Memory Score vs. MSI Score with/without Background Music",
    x = "Total Gold MSI Score",
    y = "Memory Score",
    color = "With Music (Yes or No)"
  ) +
  theme_minimal()
```

### Column 2 {width="40%"}
```{r}
#| title: Boxplot of Memory Scores Based on With/Without Background Music
#| echo: false


stats <- df %>%
  group_by(With.music.or.not) %>%
  summarise(
    Q1 = quantile(Memory.Score, 0.25),
    Q3 = quantile(Memory.Score, 0.75),
    IQR = Q3 - Q1,
    lower = max(min(Memory.Score), Q1 - 1.5 * IQR),
    upper = min(max(Memory.Score), Q3 + 1.5 * IQR),
    mean = mean(Memory.Score),
    median = median(Memory.Score),
    min = min(Memory.Score),
    max = max(Memory.Score),
    xpos = as.numeric(With.music.or.not),
    .groups = "drop"
  )

ggplot(df, aes(
  x = With.music.or.not,
  y = Memory.Score
)) +
  geom_boxplot(
    width = 0.25,
    outlier.shape = 16,
    outlier.size = 2.5,
    color = "black",
    fill = NA
  ) +
  geom_segment(
    data = stats,
    aes(x = xpos - 0.1, xend = xpos + 0.1, y = lower, yend = lower),
    linewidth = 0.9
  ) +
  geom_segment(
    data = stats,
    aes(x = xpos - 0.1, xend = xpos + 0.1, y = upper, yend = upper),
    linewidth = 0.9
  ) +
  geom_segment(
    data = stats,
    aes(x = xpos - 0.1, xend = xpos + 0.1, y = mean, yend = mean),
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  geom_text(
    data = stats,
    aes(x = xpos + 0.15, y = median, label = paste0("Median: ", round(median, 2))),
    hjust = 0,
    size = 3.5
  ) +
  geom_text(
    data = stats,
    aes(x = xpos + 0.15, y = mean, label = paste0("Mean: ", round(mean, 2))),
    hjust = 0,
    size = 3.5
  ) +
  geom_text(
    data = stats,
    aes(x = xpos + 0.15, y = lower, label = paste0("Min: ", round(min, 2))),
    hjust = 0,
    size = 3.5
  ) +
  geom_text(
    data = stats,
    aes(x = xpos + 0.15, y = upper, label = paste0("Max: ", round(max, 2))),
    hjust = 0,
    size = 3.5
  ) +
  scale_y_continuous(limits = c(0, 22), breaks = seq(0, 20, 2)) +
  labs(
    title = "Score from Memory Test by With Background Music or Not",
    x = "With Background Music or Not",
    y = "Score from Memory Test"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```

### Column 2 {width="40%"}
```{r}
#| title: How Does the Time Taken to Memorize Words Impact Test Score?
#| echo: false

# Plot
ggplot(df, aes(
    x = Time,
    y = Memory.Score,
    color = With.music.or.not
)) +
    geom_point(size = 3, alpha = 0.7) +
    geom_smooth(method = "lm", se = FALSE) +
    scale_color_manual(
        values = c(
            "No" = "#1f77b4",   # blue
            "Yes" = "#ff7f0e"   # orange
        )
    ) +
    labs(
        title = "Time (seconds) vs Memory Score by Background Music Condition",
        x = "Time (seconds)",
        y = "Memory Score",
        color = "With Music"
    ) +
    theme_minimal(base_size = 14)
```

# Our Research 

## Column 1 {Width = "50%"}

### Methodology {Height = "50%"}

Methodology


### Survey {Height = "50%"}

Survey
<audio controls>
  <source src="audiofil.mp3" type="audio/mpeg">
</audio>  <br>

## Column 2 {Width = "50%"}

### Gold MSI {Height = "50%"}

Gold MSI

### Data {Height = "50%"}

Data

# Results
## Row 1
### Column 1 {.tabset width="50%"}
#### Memory Test Scores

```{r}
#| title: Memory Test Scores dependant on Background Music and Music Proficiency
#| echo: false

ggplot(df, aes(
  x = Total.Gold.MSI.Score,
  y = Memory.Score,
  color = With.music.or.not
)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(
    values = c(
      "No" = "#1f77b4",
      "Yes" = "#ff7f0e"
    )
  ) +
  labs(
    title = "Memory Score vs. MSI Score with/without Background Music",
    x = "Total Gold MSI Score",
    y = "Memory Score",
    color = "With Music (Yes or No)"
  ) +
  theme_minimal()
```

#### Boxplot Memory Scores

```{r}
#| title: Boxplot of Memory Scores based on with/without Background Music
#| echo: false


stats <- df %>%
  group_by(With.music.or.not) %>%
  summarise(
    Q1 = quantile(Memory.Score, 0.25),
    Q3 = quantile(Memory.Score, 0.75),
    IQR = Q3 - Q1,
    lower = max(min(Memory.Score), Q1 - 1.5 * IQR),
    upper = min(max(Memory.Score), Q3 + 1.5 * IQR),
    mean = mean(Memory.Score),
    median = median(Memory.Score),
    min = min(Memory.Score),
    max = max(Memory.Score),
    xpos = as.numeric(With.music.or.not),
    .groups = "drop"
  )

ggplot(df, aes(
  x = With.music.or.not,
  y = Memory.Score
)) +
  geom_boxplot(
    width = 0.25,
    outlier.shape = 16,
    outlier.size = 2.5,
    color = "black",
    fill = NA
  ) +
  geom_segment(
    data = stats,
    aes(x = xpos - 0.1, xend = xpos + 0.1, y = lower, yend = lower),
    linewidth = 0.9
  ) +
  geom_segment(
    data = stats,
    aes(x = xpos - 0.1, xend = xpos + 0.1, y = upper, yend = upper),
    linewidth = 0.9
  ) +
  geom_segment(
    data = stats,
    aes(x = xpos - 0.1, xend = xpos + 0.1, y = mean, yend = mean),
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  geom_text(
    data = stats,
    aes(x = xpos + 0.15, y = median, label = paste0("Median: ", round(median, 2))),
    hjust = 0,
    size = 3.5
  ) +
  geom_text(
    data = stats,
    aes(x = xpos + 0.15, y = mean, label = paste0("Mean: ", round(mean, 2))),
    hjust = 0,
    size = 3.5
  ) +
  geom_text(
    data = stats,
    aes(x = xpos + 0.15, y = lower, label = paste0("Min: ", round(min, 2))),
    hjust = 0,
    size = 3.5
  ) +
  geom_text(
    data = stats,
    aes(x = xpos + 0.15, y = upper, label = paste0("Max: ", round(max, 2))),
    hjust = 0,
    size = 3.5
  ) +
  scale_y_continuous(limits = c(0, 22), breaks = seq(0, 20, 2)) +
  labs(
    title = "Score from Memory Test by With Background Music or Not",
    x = "With Background Music or Not",
    y = "Score from Memory Test"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```

#### Time Taken to Memorize 
```{r}
#| title: How does the time taken to memorize words impact test score?
#| echo: false

# Plot
ggplot(df, aes(
    x = Time,
    y = Memory.Score,
    color = With.music.or.not
)) +
    geom_point(size = 3, alpha = 0.7) +
    geom_smooth(method = "lm", se = FALSE) +
    scale_color_manual(
        values = c(
            "No" = "#1f77b4",   # blue
            "Yes" = "#ff7f0e"   # orange
        )
    ) +
    labs(
        title = "Time (seconds) vs Memory Score by Background Music Condition",
        x = "Time (seconds)",
        y = "Memory Score",
        color = "With Music"
    ) +
    theme_minimal(base_size = 14)
```
### Column 2 {.tabset width="50%"}
#### Hypotheses 
explain hypotheses 

#### GoldMSI
explain goldmsi


# Appendix 

## Everything {.tabset}

### Other Graphs {.tabset}

#### Heatmap
Insert graphs

#### etc
Insert graphs


### Article Reviews {.tabset}

#### Review 1
Insert Article Reviews here

#### Review 2 
Insert Article Reviews here

#### Review 3
Insert Article Reviews here

#### Review 4 
Insert Article Reviews here

#### Review 5 
Insert Article Reviews here

### Resources 

Insert resources or bibliography or whatever


