---
title: Music Cognition and Memory
format:
  dashboard:
    theme: 
    - superhero
    - styles.scss
    orientation: columns
---

```{r setup}
#| include: false
library(tidyverse)
```

# Introduction

##  {.sidebar}

For our research on everyday music listening we decided to investigate the question

How does passive music listening affect performance for people of different levels of music cognition?

We think that it is worth investigating the possibility that people with a high knowledge of music theory might get more or less distracted/affected by the presence of background music when performing a memory task.

To do this we presented all of our subjects with the exact same memory task, except half of them (selected randomly) had to perform the tasks

## Column 1 {width="60%"}

```{r}
#| title: Chart 1
ggplot()
```

## Column 2 {width="40%"}

### Row 1

This is more text. Below it, you'll see another row with a tabset.

### Row 2 {.tabset}

```{r}
#| title: Chart 2
ggplot()
```

```{r}
#| title: Chart 3
ggplot()
```

# The Data


# Methodology

# Extras {}

